\coloc: Optimize Spatial Visualization Following Gestalt Principles

### 2023-03-17 Manuscript/manuscript.tex
Edit Abstract

%% ABSTRACT ============================================================================================
\section*{Abstract}
\subsection*{Summary}
Colocalization is highly interested research questions in the analysis of spatial omics data. While many sophisticated computational models are proposed to infer the colocalization relationship between genes or proteins in the spatial context, there completely lack of intuitive spatial visualization, making interpretation extremely challenging. The biggest challenge for colocalization visualization is two sources of orthogonal information, spot colocalization and spot membership (such as spatial domain) the same spatial plot. To address this issue, we developed \coloc to display the colocalization while maintaining an intuitive representation of the surrounding spatial domain. \coloc leverages color contrast optical illusion to display spatial domains to create shadows representing different spatial domain while using different spots to denote colocalization status.

\subsection*{Availability and implementation}
The R package \coloc is freely available at Github (\url{https://github.com/boyiguo1/VisColoc}).
% and Zenodo (\fixme{add link}).

@@

The recent technology breakthrough in spatial profiling enables characterizing the diversity of molecules, e.g. DNA, RNA, and proteins, in the spatial context\cite{moffitt_2022}, generating novel spatially-resolved data that requires highly tailored \textit{in situ} visualizations\cite{dries_2021, Lewis_2021, odonoghue_2021}. Current \textit{in situ} visualizations, as opposed to embedding visualizations (e.g. t-SNE and UMAP) that project information to some mathematical space, visualize molecular information in the original cellular location, creating spatial maps that color-code a single variable (e.g., molecule expression, cell type, or spatial domain). This approach lacks the ability to simultaneously exhibit multiple variables, possibly from disparate data domains (such as expression domain and spatial domain) or modalities (such as transcriptomics and proteomics), creating cognitive gaps to interpret findings regarding their (micro-) environment. For example, when studying differential expression in different spatial domains (commonly defined by tissue architecture, e.g. cancerous regions, cortical layers), two spatial maps are displayed side-by-side, one for gene expression and one for spatial domains, creating significant challenges to associate these two information together. While software-based interactive visualizations\cite{sriworarat_2023} or 3-D visualizations have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as prints. Developing a static spatial visualization that enables the simultaneous display of multiple dimensions of information is crucial for spatially-resolved data research.

@@

\section*{Use Case 2: Multi-dimensional UMAP}
The application of the proposed framework doesn't limit to in-situ visualizations of spatially resolved data. It can be broadly applied to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and t-SNE. To demonstrate, we apply the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, \texttt{schex} proposed to apply hexagonal binning strategy to unbiasedly display the gene expression.\cite{freytag_2020} In addition, color-coded convex hulls were used to annotate different clusters of cells (Figure \label{fig:visual}). Nevertheless, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we integrate the proposed visualization to the \texttt{schex} by replacing convex hulls with color-coded hexagons boundaries to avoid possible membership confusion. The visual improvement is easily implemented without any modification of \texttt{schex} thanks to the Grammar of Graphics \cite{wilkinson_2012} standard.

\section*{Summary}

While \coloc is to design to display colocalization in their spatial context, the underlying principle can be broadly applied to address visualizaiton challenges in informaiton rich visualizations. The contrast color optical illusion provides allows to add another dimension of information in any spatial visualization, which can be further used to display multi-omics presence, greatly improve the interpretability.

In this manuscript, we demonstrate the utility of the proposed spatial visualization using spot-based spatial resolved data, nevertheless, this method can be applied to image-based spatial resolved data by applying \texttt{schex} to avoid overplotting commonly for image-based spatial resolved data, similar to our second UMAP example where the first step is to reduce done the dimensiona

our programming provides flexibility by following the grammar of graphs, especially ggplot2

translating the findings in vision science to dat science to address many of the current challenges in visualization.

>>>>

%% ABSTRACT ============================================================================================
\section*{Abstract}
\subsection*{Summary}
Visualization is an indispensable component of data analysis and receives growing recognition as essential. Nevertheless, current methods and
tools are inadequate to visualize the novel spatially-resolved data, limiting the ability to derive new insights and discoveries. Specifically, there are no multi-dimensional \textit{in-situ} visualization that simultaneously displays multiple variables in a 2D spatial map. To address this problem, we introduce gestalt principles to the design of spatial maps, layering aesthetics to display multiple variables. The proposed visualization can be broadly applied to spatially resolved data and 2D embedding methods. We provide an open-source R package \coloc, which seamlessly integrates into the state-of-the-art spatial omics toolboxes.

\subsection*{Availability and implementation}
The R package \coloc is freely available at Github (\url{https://github.com/boyiguo1/escheR}).
% and Zenodo (\fixme{add link}).

@@

The recent technology breakthrough in spatial profiling enables characterizing the diversity of molecules, e.g. DNA, RNA, and proteins, in the spatial context\cite{moffitt_2022}, generating novel spatially-resolved data that requires highly tailored \textit{in situ} visualizations\cite{dries_2021, Lewis_2021, odonoghue_2021}. Current \textit{in situ} visualizations, as opposed to embedding visualizations (e.g. t-SNE \cite{hinton_2002} and UMAP \cite{becht_2019}) that project information to some mathematical space, visualize molecular information in the original cellular location, creating spatial maps that color-code a single variable (e.g., molecule expression, cell type, or spatial domain). This approach lacks the ability to simultaneously exhibit multiple variables, possibly from disparate data domains (such as expression domain and spatial domain) or modalities (such as transcriptomics and proteomics), creating cognitive gaps to interpret findings regarding their (micro-) environment. For example, when studying differential expression in different spatial domains (commonly defined by tissue architecture, e.g. cancerous regions, cortical layers), two spatial maps are displayed side-by-side, one for gene expression and one for spatial domains, creating significant challenges to associate these two information together. While software-based interactive visualizations\cite{sriworarat_2023} or 3-D visualizations have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as prints. Developing a static spatial visualization that enables the simultaneous display of multiple dimensions of information is crucial for spatially-resolved data research.

@@

\section*{Use Case 2: Multi-dimensional UMAP}
The application of the proposed framework doesn't limit to in-situ visualizations of spatially resolved data. It can be broadly applied to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and t-SNE. To demonstrate, we apply the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, \texttt{schex} proposed to apply hexagonal binning strategy to unbiasedly display the gene expression.\cite{freytag_2020} In addition, color-coded convex hulls were used to annotate different clusters of cells (Figure \label{fig:visual}). Nevertheless, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we integrate the proposed visualization to \texttt{schex} by replacing convex hulls with color-coded hexagons boundaries  (Figure \label{fig:visual}) to avoid possible membership confusion. The visual improvement is easily implemented without any modification of \texttt{schex} thanks to the Grammar of Graphics \cite{wilkinson_2012} standard.

\vspace{0.45in}
In summary, we propose an innovative multi-dimensional spatial visualization to simultaneously display multiple variables in a 2D coordinate system. Specifically, our design leverages the gestalt principle from visual perception to introduce other dimensions of a spatial map by iterative layering aesthetics. Developed upon \texttt{ggplot2}, we provide an open-source R package \coloc that is seamlessly compatible with popular spatial resolved data analysis toolboxes. The proposed visualization has broad applications in visualizing both spot-based and image-based spatial resolved data, spatial multi-omics as well as other spatial-based data. The proposed visualization innovation highlights the importance and potential of translating theories in vision science to data science, addressing long-standing challenges in bioinformatics data visualization.

% While \coloc is to design to display colocalization in their spatial context, the underlying principle can be broadly applied to address visualizaiton challenges in informaiton rich visualizations. The contrast color optical illusion provides allows to add another dimension of information in any spatial visualization, which can be further used to display multi-omics presence, greatly improve the interpretability.

% our programming provides flexibility by following the grammar of graphs, especially ggplot2
